Refinement of Parallel Algorithms down to LLVM

### 2022-05-12 papers/ITP2022/main.tex
renaming files

\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

@@

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\noindent Formalization available at: \url{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}

\subsection{Overview}\label{sec:intro_overview}
This paper is based on the Isabelle Refinement Framework (IRF), a continuing effort
to verify efficient implementations of complex algorithms, using stepwise refinement techniques.
Figure~\ref{fig:IRF-overview} displays the components of the Isabelle Refinement Framework.

@@

This basic program logic and VCG can already be used to verify simple low-level algorithms and data structures,
like dynamic arrays and linked lists. More complex developments typically use a stepwise refinement approach,
starting at purely functional programs modelled in a nondeterminism-error (NE) monad~\cite{LaTu12}.
A semi-automatic refinement procedure (Sepref~\cite{La15,La19-llvm}) translates from the purely functional
code to imperative code, refining abstract functional data types to concrete imperative ones.
In Section~\ref{sec:refine_parallel}, we describe our extensions to support refinement to parallel executions,
and a fine-grained tracking of pointer equalities, required to parallelize computations that work on
disjoint parts of the same array.

@@

  \subsection{State-Nondeterminism-Error Monad with Access Reports}\label{sec:monad}
  We define the underlying monad in two steps. We start with a nondeterminism-error monad,
  and then lift it to a state monad and add access reports.
  Defining a nondeterminism-error monad is standard:
  \begin{lstlisting}
    'a neM == spec ("'a => bool") | fail
    return x == spec (%r. r=x)
    bind fail f == fail
    bind (spec P) f == if ?x. P x & f x = fail then fail
                      else spec (%r. ?x Q. P x & f x = spec Q & Q r)
  \end{lstlisting}
  A program either fails, or yields a possible set of results (\is{spec P}), described by its characteristic function \is{P}.
  The \is{return} operation yields exactly one result, and \is{bind} combines all possible results,
  failing if there is a possibility to fail.

  Now assume that we have a state (memory) type \is{'\mu}, and an access report type \is{'\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == '\mu => ('x \x '\rho \x '\mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, and adds up the access reports.

  Typically, the access report will contain read and written addresses, such that we can detect a data race.
  Moreover, if parallel executions can allocate memory, we must detect those
  executions where the memory manager allocated the same block in both parallel strands.
  As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
  Let \is{norace :: '\rho => '\rho => bool} and \is{feasible :: '\rho => '\rho => bool} be symmetric predicates,
  and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
  compose two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
  \begin{lstlisting}
    (m_1 || m_2) \mu ==
      (x_1,(r_1,\mu_1)) <- m_1 \mu; (x_2,(r_2,\mu_2)) <- m_2 \mu;  --"\rightcomment{execute both strands}"
      assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
      assert norace \rho_1 \rho_2; --"\rightcomment{fail on data race}"
      return$_{ne}$ ((x_1,x_2), combine (\rho_1,\mu_1) (\rho_2,\mu_2))  --"\rightcomment{combine results}"

@@

  To access and modify memory, we define the functions \is{valid}, \is{get}, and \is{put}:
  \begin{lstlisting}
    valid \mu (ADDR b i) == is_alloc (\mu b) & i<|vals (\mu b)|
    get \mu (ADDR b i) == vals (\mu b) ! i
    put \mu (ADDR b i) x == \mu(b := BLOCK ((vals (\mu b))[i:=x]))
  \end{lstlisting}
  where \is{|xs|} is the length of list \is{xs}, \is{xs!i} returns the \is$i$th element of list \is$xs$,
  and \is{xs[i:=x]} replaces the \is$i$th element of \is{xs} by \is{x}.

@@

    norace (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) ==
      let m_1 = w_1 \<union> { ADDR b i. b \<in> a_1 \<union> f_1 } in
      let m_2 = w_2 \<union> { ADDR b i. b \<in> a_2 \<union> f_2 } in
        (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
  \end{lstlisting}
  The invariant for \is{M} states that blocks transition only from fresh to allocated to free,
  allocated blocks never change their size, and the access report matches the observable state change (\is{consistent}).
  It also states, that for each finite set of blocks $B$, there is an execution that does not allocate blocks from $B$.
  The latter is required to show that we always find feasible parallel executions:
  \begin{lstlisting}
    invar$_M$ c == \<forall>\mu P. c \mu = spec P ==>
      (\<forall>x \rho \mu'. P (x,\rho,\mu') ==> consistent \mu \rho \mu')
    & (\<forall>B. finite B ==> (\<exists>x \rho \mu'. P (x,\rho,\mu') & \rho.a \inter B = {} ))
  \end{lstlisting}
  The combine function joins the access reports and memories,
  preferring allocated over fresh, and freed over allocated memory. When joining two allocated blocks, the written addresses
  from the access report are used to join the blocks. We skip the rather technical definition of combine, and just state the
  relevant properties: Let \is{\rho_1=(r_1,w_1,a_1,f_1)} and \is{\rho_2=(r_2,w_2,a_2,f_2)} be feasible and race free access reports,
  and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}, and \is{addr} a valid address in \is{\mu'}. Then
  \begin{lstlisting}
    \mu' b = FRESH <--> if b \in a_2 \union f_2 then \mu_2 b = FRESH else \mu_1 b = FRESH
    \mu' b = FREED <--> if b \in a_2 \union f_2 then \mu_2 b = FREED else \mu_1 b = FREED
    is_alloc (\mu' b) <--> if b \in b \in a_2 then is_alloc (\mu_2 b) else is_alloc (\mu_1 b)
    get \mu' addr = if addr \in w_2 \or addr.bidx \in a_2 \union f_2 then get \mu_2 addr else get \mu_1 addr
  \end{lstlisting}

@@

    \alpha$_m$ \mu addr == if valid \mu addr then Some (get \mu addr) else 0
    \alpha$_b$ \mu b == if is_alloc (\mu b) then TRIV (|vals (\mu b)|) else 0
  \end{lstlisting}
  An abstract memory \is{\alpha \mu} consists of two parts:
  \is{\alpha$_m$ \mu} is a map from addresses to the values stored there. It is used to reason about load and store operations.
  \is{\alpha$_b$ \mu} is a map from block indexes to the sizes of the corresponding blocks.
    It is used to ensure that one owns all addresses of a block when freeing it.

  We continue to define a separation logic: assertions are predicates over separation algebra elements.
  The basic connectives are defined as follows:
  \begin{lstlisting}
    false a == False     true a == True     \<box> a == a=0
    (P*Q) a == \<exists> a_1 a_2. a_1 # a_2 \and a = a_1 + a_2 \and P a_1 \and Q a_2
  \end{lstlisting}
  That is, the assertion \is{false} never holds and the assertion \is{true} holds for all abstract memories.
  The empty assertion \is{\<box>} holds for the zero memory, and the separating conjunction \is{P*Q}
  holds if the memory can be split into two disjoint parts, such that \is{P} holds for one,
  and \is{Q} holds for the other part. The lifting assertion \is{\<up>\<phi>} holds iff the Boolean value \is{\<phi>} is true:
  \begin{lstlisting}
    \<up>\<phi> == if \<phi> then \<box> else false
  \end{lstlisting}
  It does not depend on the memory, and is used to lift plain logical statements into separation logic.
  When clear from the context, we omit the \is{\<up>}-symbol, and just mix plain statements with separation logic assertions.
%   , e.g., \is{array xs p * i<|xs|}

@@

  Equipped with a weakest precondition with access restrictions, we define a Hoare-triple:
  \begin{lstlisting}
    ABS amf P \mu == \<exists>am. am ## amf \and \<alpha> \mu = am+amf \and P am

@@

  \subsection{Verification Condition Generator}
  The verification condition generator is implemented as a proof tactic that works on subgoals of the form:
  \begin{lstlisting}
    ABS amf P \mu \and ... ==> wpf amf c Q \mu
  \end{lstlisting}
  The tactic is guided by the syntax of the command \is{c}.
  Basic monad combinators are broken down using the following rules:
  \begin{lstlisting}
    Q r \mu ==> wpf amf (return r) \mu
    wpf amf m (%x. wpf amf (f x) Q) \mu ==> wpf amf ({x <- m; f x}) Q \mu
  \end{lstlisting}
  For other instructions and user defined functions, the VCG expects a Hoare-triple to be already proved.
  It then uses the following rule:
  \begin{lstlisting}
    ht P c Q \and ABS amf P' \mu   --"\rightcomment{match Hoare triple and current state}"
    \and P' |- P*F --"\rightcomment{infer frame}"
    \and (!!r \mu. ABS amf (Q r ** F) \mu ==> Q' r \mu) --"\rightcomment{continue with postcondition}"
    ==> wpf amf c Q' \mu
  \end{lstlisting}
  To process a command \is{c}, the first assumption is instantiated with the Hoare-triple for \is{c},
  and the second assumption with the assertion \is{P'} for the current state.
  Then, a simple syntactic heuristics infers a frame \is{F} and proves that the
  current assertion \is{P'} entails the required precondition \is{P} and the frame.
  Finally, verification condition generation continues with the postcondition \is{Q} and the frame as current assertion.

@@

  For memory operations, we prove:
  \begin{lstlisting}
  |= {n~=0} ll_malloc TYPE(\<alpha>) n {%p. range {0..<n} (%_. init) p * b_tag n p}
  |= {range {0..<n} xs p * b_tag n p} ll_free p {%_. \<box>}
  |= {pto x p} ll_load p {%r. r=x * pto x p}
  |= {pto xx p} ll_store x p {%_. pto x p}
  \end{lstlisting}
  Here \is{b_tag n p} asserts that \is{p} points to the beginning of a block of size \is$n$, and
  \is{range I f p} describes that for all $i\in I$, $p+i$ points to value $f~i$.
  Intuitively, \is{ll_malloc} creates a block of size $n$, initialized with the default \is{init} value,
  and a tag. If one possesses both, the whole block and the tag, it can be deallocated by free.
  The rules for load and store are straightforward, where \is{pto x p} describes that $p$ points to value $x$.

@@

  \subsection{Abstract Programs}
  Abstract programs are shallowly embedded into the nondeterminism error monad \is{'a ne} (cf.\ Section~\ref{sec:monad}).
  They are purely functional, not modifying memory, or differentiating between sequential and parallel execution.
  We define a \emph{refinement ordering} on \is{ne}:
  \begin{lstlisting}
    spec P <= spec Q == \<forall>x. P x ==> Q x         fail $\not\le$ spec Q         m <= fail
  \end{lstlisting}
  Intuitively, \is{m_1 <= m_2} means
  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
  Note that \is$<=$ is a complete lattice, with top element \is$fail$.

@@

  For a detailed description on using the \is{ne} monad for stepwise refinement based program verification, we refer the reader to~\cite{LaTu12}.

  \subsection{The Sepref Tool}
  The Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{ne}-monad,
  keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
  The synthesis is automatic, but requires use annotations to the abstract program.

  The main concept of the Sepref tool is refinement between an abstract program \is{c} in the \is{ne} monad,
  and a concrete program \is{c$_\dagger$} in the \is{M} monad, as expressed by the \is{hnr}-predicate:
  \begin{lstlisting}
    "hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ==
        c ~= fail ==> ht \<Gamma> c$_\dagger$ (%x. \<exists>x$_\dagger$. \<Gamma>' * R x x$_\dagger$ * \<up>(return x <= c \and CP x$_\dagger$))"
  \end{lstlisting}
  That is, either the abstract program \is{c} fails,
  or for a memory described by assertion \is{\<Gamma>},
  the LLVM program \is{c$_\dagger$} succeeds with \is{r$_\dagger$},
  such that the new memory is described by \is{\<Gamma>' * R x x$_\dagger$},
  for a possible result \is{x} of the abstract program \is{c}. Moreover, the predicate \is{CP} holds for the concrete result.
  Note that \is{hnr} trivially holds for a failing abstract program. This makes sense, as we prove that
  the abstract program does not fail anyway. Moreover it allows us to assume that assertions actually hold during the refinement proof:
  \begin{lstlisting}
    ( \<phi> ==> hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ) ==> hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP (assert \<phi>; c)
  \end{lstlisting}

@@

  Given refinement assertions for the parameters, and \is{hnr}-rules for all operations in a program,
  the Sepref tool automatically synthesizes an LLVM program from an abstract \is{ne} program.
  The tool tries to automatically discharge additional proof obligations, typically arising from
  translating arithmetic operations from unbounded numbers to fixed width numbers. Where automatic proof fails,
  the user has to add assertions to the abstract program to help the proof.
  The main difference of our tool wrt.\ the existing Sepref tool~\cite{La19-llvm} is the additional
  condition (\is{CP}) on the concrete result, which is used to track pointer equalities.
  We have added a heuristics to automatically synthesize and discharge these equalities.

@@

  We resolve such closures over the comparison function manually: using Isabelle's locale mechanism~\cite{KWP99},
  we parametrize over the comparison function. Moreover, we thread through an extra parameter for the data
  captured by the closure:
  \begin{lstlisting}
    locale pcmp =
      fixes lt :: 'p => 'e => 'e => bool and lt$_\dagger$ :: 'p$_\dagger$ => 'e$_\dagger$ => 'e$_\dagger$ => bool
        and par$_A$ :: 'p => 'p$_\dagger$ => assn and elem$_A$ :: 'e => 'e$_\dagger$ => assn
      assumes \<forall>p. weak_ordering (lt p)
      assumes hnr (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi) (lt$_\dagger$ pi ai bi)
        (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True) (lt p a b)
  \end{lstlisting}
  This defines a context in which we have an abstract compare function \is{lt} for the abstract elements of type \is{'e}.
  It takes an extra parameter of type \is{'p} (e.g.\ the list $xs$), and forms a weak ordering\footnote{
    A weak ordering is induced by a mapping of the elements into a total ordering. It is the standard prerequisite for sorting algorithms in C++~\cite{Josu12}.
  }. Moreover, we have a concrete implementation \is{lt$_\dagger$} or the compare function, wrt.\ the refinement assertions
  \is{par$_A$} for the parameter and \is{elem$_A$} for the elements.

>>>>

\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{color}
\usepackage[noend]{algpseudocode}
\usepackage{caption}
\usepackage[font=scriptsize]{subcaption}
\usepackage{hyperref}
\usepackage{float}
% \usepackage{wrapfig}

@@

\supplementdetails[subcategory={Isabelle Formalization}]{Software}{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}

@@

% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \EventEditors{John Q. Open and Joan R. Access}
% \EventNoEds{2}
% \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
% \EventShortTitle{CVIT 2016}
% \EventAcronym{CVIT}
% \EventYear{2016}
% \EventDate{December 24--27, 2016}
% \EventLocation{Little Whinging, United Kingdom}
% \EventLogo{}
% \SeriesVolume{42}
% \ArticleNo{23}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\EventEditors{June Andronick and Leonardo de Moura}
\EventNoEds{2}
\EventLongTitle{13th International Conference on Interactive Theorem Proving (ITP 2022)}
\EventShortTitle{ITP 2022}
\EventAcronym{ITP}
\EventYear{2022}
\EventDate{August 7--10, 2022}
\EventLocation{Haifa, Israel}
\EventLogo{}
\SeriesVolume{237}
\ArticleNo{12}

@@

% \noindent The formalization is available at \url{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}.

\subsection{Overview}\label{sec:intro_overview}
This paper is based on the Isabelle Refinement Framework, a continuing effort
to verify efficient implementations of complex algorithms, using stepwise refinement techniques.
Figure~\ref{fig:IRF-overview} displays the components of the Isabelle Refinement Framework.

@@

At the level of the program logic and VCG, our framework can already be used to verify simple low-level algorithms and data structures,
like dynamic arrays and linked lists. More complex developments typically use a stepwise refinement approach,
starting at purely functional programs modelled in a nondeterminism-error (NE) monad~\cite{LaTu12}.
A semi-automatic refinement procedure (Sepref~\cite{La15,La19-llvm}) translates from the purely functional
code to imperative code, refining abstract functional data types to concrete imperative ones.
In Section~\ref{sec:refine_parallel}, we describe our extensions to support refinement to parallel executions,
and a fine-grained tracking of pointer equalities, required to parallelize computations that work on
disjoint parts of the same array.

@@

  \subsection{State-Nondeterminism-Error Monad with Access Reports}\label{sec:monad}
  We define the underlying monad in two steps. We start with a nondeterminism-error monad,
  and then lift it to a state monad and add access reports.
  Defining a nondeterminism-error monad is straightforward in Isabelle/HOL:
  \begin{lstlisting}
    'a neM == spec ("'a => bool") | fail
    return x == spec (%r. r=x)
    bind fail f == fail
    bind (spec P) f == if ?x. P x & f x = fail then fail
                      else spec (%r. ?x Q. P x & f x = spec Q & Q r)
  \end{lstlisting}
  A program either fails, or yields a possible set of results (\is{spec P}), described by its characteristic function \is{P}.
  The \is{return} operation yields exactly one result, and \is{bind} combines all possible results,
  failing if there is a possibility to fail.

  Now assume that we have a state (memory) type \is{'\mu}, and an access report type \is{'\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == '\mu => ('x \x '\rho \x '\mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, threading through the state $\mu$, and adding up the access reports \is$r_1$ and \is$r_2$.

  Typically, the access report will contain read and written addresses, such that data races can be detected.
  Moreover, if parallel executions can allocate memory, we must detect those
  executions where the memory manager allocated the same block in both parallel strands.
  As we assume a thread safe memory manager, those \emph{infeasible} executions can safely be ignored.
  Let \is{norace :: '\rho => '\rho => bool} and \is{feasible :: '\rho => '\rho => bool} be symmetric predicates,
  and let \is{combine :: ('\rho \x '\mu) => ('\rho \x '\mu) => ('\rho \x '\mu) } be a commutative operator to
  compose two pairs of access reports and states. Then, we define a parallel composition operator for \is{M}:
  \begin{lstlisting}
    (m_1 || m_2) \mu ==
      (x_1,r_1,\mu_1) <- m_1 \mu; (x_2,r_2,\mu_2) <- m_2 \mu;  --"\rightcomment{execute both strands}"
      assume feasible \rho_1 \rho_2;  --"\rightcomment{ignore infeasible combinations}"
      assert norace \rho_1 \rho_2; --"\rightcomment{fail on data race}"
      return$_{ne}$ ((x_1,x_2), combine (\rho_1,\mu_1) (\rho_2,\mu_2))  --"\rightcomment{combine results}"

@@

  To access and modify memory, we define the functions \is{valid}, \is{get}, and \is{put}:
  \begin{lstlisting}
    valid \mu (ADDR b i) == is_alloc (\mu b) & i<|vals (\mu b)|
    get \mu (ADDR b i) == vals (\mu b) ! i
    put \mu (ADDR b i) x == \mu(b := ALLOC ((vals (\mu b))[i:=x]))
  \end{lstlisting}
  where \is{|xs|} is the length of list \is{xs}, \is{xs!i} returns the \is$i$th element of list \is$xs$,
  and \is{xs[i:=x]} replaces the \is$i$th element of \is{xs} by \is{x}.

@@

    norace (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) ==
      let m_1 = w_1 \<union> { ADDR b i. b \<in> a_1 \<union> f_1 } in
      let m_2 = w_2 \<union> { ADDR b i. b \<in> a_2 \<union> f_2 } in
        (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
  \end{lstlisting}
  The invariant for \is{M} states that blocks transition only from fresh to allocated to free,
  allocated blocks never change their size, and the access report matches the observable state change (\is{consistent}).
  It also states, that for each finite set of blocks $B$, there is an execution that does not allocate blocks from $B$.
  The latter is required to show that we always find feasible parallel executions:
  \begin{lstlisting}
    invar$_M$ c == \<forall>\mu P. c \mu = spec P ==>
      (\<forall>x \rho \mu'. P (x,\rho,\mu') ==> consistent \mu \rho \mu')
    & (\<forall>B. finite B ==> (\<exists>x \rho \mu'. P (x,\rho,\mu') & \rho.a \inter B = {} ))
  \end{lstlisting}
  The combine function joins the access reports and memories,
  preferring allocated over fresh, and freed over allocated memory. When joining two allocated blocks, the written addresses
  from the access report are used to join the blocks. We skip the rather technical definition of combine, and just state the
  relevant properties: Let \is{\rho_1=(r_1,w_1,a_1,f_1)} and \is{\rho_2=(r_2,w_2,a_2,f_2)} be feasible and race free access reports,
  and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}, and \is{addr} a valid address in \is{\mu'}. Then
  \begin{lstlisting}
    (1)  \mu' b = FRESH <--> \mu b = FRESH \and b ~: a_1 \union a_2
    (2)  is_alloc (\mu' b) <--> (is_alloc (\mu b) \or b \<in> a_1 \union a_2) \and b ~: f_1 \union f_2
    (3)  \mu' b = FREED <--> \mu b = FREED \or b \in f_1 \union f_2

    (4)  a \in w_1 \or b \in a_1 ==> get_addr \mu' a = get_addr \mu_1 a
    (5)  a \in w_2 \or b \in a_2 ==> get_addr \mu' a = get_addr \mu_2 a
    (6)  a ~: w_1\<union>w_2 \or b ~: a_1\<union>a_2 ==> get_addr \mu' a = get_addr \mu a
  \end{lstlisting}
  The properties (1)--(3) define the state of blocks in the combined memory:
  a fresh block in \is{\mu'} was fresh already in \is{\mu}, and has not been allocated (1);
  an allocated block was already allocated or has been allocated, but has not been freed (2); and a freed block was already freed, or has been freed (3).
  The properties (4)--(6) define the content: addresses written or allocated in the first or second execution get their content from \is{\mu_1} (4) or \is{\mu_2} (5) respectively.
  Addresses not written or allocated at all keep their original content (6).

@@

    \alpha$_m$ \mu addr == if valid \mu addr then Some (get \mu addr) else 0
    \alpha$_b$ \mu b == if is_alloc (\mu b) then Some (|vals (\mu b)|) else 0
  \end{lstlisting}
  An abstract memory \is{\alpha \mu} consists of two parts:
  \is{\alpha$_m$ \mu} is a map from addresses to the values stored there. It is used to reason about load and store operations.
  \is{\alpha$_b$ \mu} is a map from block indexes to the sizes of the corresponding blocks.
    It is used to ensure that one owns all addresses of a block when freeing it.

  We continue to define a separation logic: assertions are predicates over separation algebra elements.
  The basic connectives are defined as follows:
  \begin{lstlisting}
    false a == False     true a == True     \<box> a == a=0
    (P*Q) a == \<exists> a_1 a_2. a_1 # a_2 \and a = a_1 + a_2 \and P a_1 \and Q a_2
  \end{lstlisting}
  That is, the assertion \is{false} never holds and the assertion \is{true} holds for all abstract memories.
  The empty assertion \is{\<box>} holds for the zero memory, and the separating conjunction \is{P*Q}
  holds if the memory can be split into two disjoint parts, such that \is{P} holds for one,
  and \is{Q} holds for the other part. The lifting assertion \is{\<up>\<phi>} holds iff the Boolean value \is{\<phi>} is true:
  \begin{lstlisting}
    \<up>\<phi> == if \<phi> then \<box> else false
  \end{lstlisting}
  It is used to lift plain logical statements into separation logic assertions owning no memory.
  When clear from the context, we omit the \is{\<up>}-symbol, and just mix plain statements with separation logic assertions.
%   , e.g., \is{array xs p * i<|xs|}

@@

  Equipped with a weakest precondition with access restrictions, we define a Hoare-triple:
  \begin{lstlisting}
    ABS amf P \mu == \<exists>am. am # amf \and \<alpha> \mu = am+amf \and P am

@@

  \subsection{Verification Condition Generator}
  The verification condition generator is implemented as a proof tactic that works on subgoals of the form:
  \begin{lstlisting}
    ABS amf P \mu \and ... ==> wpf amf c Q \mu
  \end{lstlisting}
  The tactic is guided by the syntax of the command \is{c}.
  Basic monad combinators are broken down using the following rules:
  \begin{lstlisting}
    Q r \mu ==> wpf amf (return r) Q \mu
    wpf amf m (%x. wpf amf (f x) Q) \mu ==> wpf amf ({x <- m; f x}) Q \mu
  \end{lstlisting}
  For other instructions and user defined functions, the VCG expects a Hoare-triple to be already proved.
  It then uses the following rule:
  \begin{lstlisting}
    ht P c Q \and ABS amf P' \mu   --"\rightcomment{match Hoare triple and current state}"
    \and P' |- P*F --"\rightcomment{infer frame}"
    \and (!!r \mu. ABS amf (Q r * F) \mu ==> Q' r \mu) --"\rightcomment{continue with postcondition}"
    ==> wpf amf c Q' \mu
  \end{lstlisting}
  To process a command \is{c}, the first assumption is instantiated with the Hoare-triple for \is{c},
  and the second assumption with the assertion \is{P'} for the current state.
  Then, a simple syntactic heuristics infers a frame \is{F} and proves that the
  current assertion \is{P'} entails the required precondition \is{P} and the frame.
  Finally, verification condition generation continues with the postcondition \is{Q} and the frame as current assertion.

@@

  For memory operations, we prove:
  \begin{lstlisting}
  |= {n~=0} ll_malloc TYPE(\<alpha>) n {%p. range {0..<n} (%_. init) p * b_tag n p}
  |= {range {0..<n} xs p * b_tag n p} ll_free p {%_. \<box>}
  |= {pto x p} ll_load p {%r. r=x * pto x p}
  |= {pto y p} ll_store x p {%_. pto x p}
  \end{lstlisting}
  Here \is{b_tag n p} asserts that \is{p} points to the beginning of a block of size \is$n$, and
  \is{range I f p} describes that for all $i\in I$, $p+i$ points to value $f~i$.
  Intuitively, \is{ll_malloc} creates a block of size $n$, initialized with the default \is{init} value,
  and a tag. If one possesses both, the whole block and the tag, it can be deallocated by free.
  The rules for load and store are straightforward, where \is{pto x p} describes that $p$ points to value $x$.

@@

  \subsection{Abstract Programs}
  Abstract programs are shallowly embedded into the nondeterminism error monad \is{'a neM} (cf.\ Section~\ref{sec:monad}).
  They are purely functional, not modifying memory, or differentiating between sequential and parallel execution.
  We define a \emph{refinement ordering} on \is{neM}:
  \begin{lstlisting}
    spec P <= spec Q == \<forall>x. P x ==> Q x         fail $\not\le$ spec Q         m <= fail
  \end{lstlisting}
  Intuitively, \is{m_1 <= m_2} means
  that \is{m_1} returns fewer possible results than \is{m_2}, and may only fail if \is{m_2} may fail.
  Note that \is$<=$ is a complete lattice, with top element \is$fail$.

@@

  For a detailed description on using the \is{ne}-monad for stepwise refinement based program verification, we refer the reader to~\cite{LaTu12}.

  \subsection{The Sepref Tool}
  The Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{ne}-monad,
  keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
  The synthesis is automatic, but requires annotations to the abstract program.

  The main concept of the Sepref tool is refinement between an abstract program \is{c} in the \is{ne}-monad,
  and a concrete program \is{c$_\dagger$} in the \is{M} monad, as expressed by the \is{hnr}-predicate:
  \begin{lstlisting}
    "hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ==
        c ~= fail ==> ht \<Gamma> c$_\dagger$ (%x$_\dagger$. \<exists>x. \<Gamma>' * R x x$_\dagger$ * \<up>(return x <= c \and CP x$_\dagger$))"
  \end{lstlisting}
  That is, either the abstract program \is{c} fails,
  or for a memory described by assertion \is{\<Gamma>},
  the LLVM program \is{c$_\dagger$} succeeds with \is{x$_\dagger$},
  such that the new memory is described by \is{\<Gamma>' * R x x$_\dagger$},
  for a possible result \is{x} of the abstract program \is{c}. Moreover, the predicate \is{CP} holds for the concrete result.
  Note that \is{hnr} trivially holds for a failing abstract program. This makes sense, as we prove that
  the abstract program does not fail anyway. Moreover it allows us to assume that assertions actually hold during the refinement proof:
  \begin{lstlisting}
    ( \<phi> ==> hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP c ) ==> hnr \<Gamma> c$_\dagger$ \<Gamma>' R CP (assert \<phi>; c)
  \end{lstlisting}

@@

  Given refinement assertions for the parameters, and \is{hnr}-rules for all operations in a program,
  the Sepref tool automatically synthesizes an LLVM program from an abstract \is{neM} program.
  The tool tries to automatically discharge additional proof obligations, typically arising from
  translating arithmetic operations from unbounded numbers to fixed width numbers. Where automatic proof fails,
  the user has to add assertions to the abstract program to help the proof.
  The main difference of our tool wrt.\ the existing Sepref tool~\cite{La19-llvm} is the additional
  condition (\is{CP}) on the concrete result, which is used to track pointer equalities.
  We have added a heuristics to automatically synthesize and discharge these equalities.

@@

  We resolve such closures over the comparison function manually: using Isabelle's locale mechanism~\cite{KWP99},
  we parametrize over the comparison function. Moreover, we thread through an extra parameter for the data
  captured by the closure:
  \begin{lstlisting}
    locale pcmp =
      fixes lt :: 'p => 'e => 'e => bool and lt$_\dagger$ :: 'p$_\dagger$ => 'e$_\dagger$ => 'e$_\dagger$ => bool
        and par$_A$ :: 'p => 'p$_\dagger$ => assn and elem$_A$ :: 'e => 'e$_\dagger$ => assn
      assumes \<forall>p. weak_ordering (lt p)
      assumes hnr (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi) (lt$_\dagger$ pi ai bi)
        (par$_A$ p pi * elem$_A$ a ai * elem$_A$ b bi)  (bool$_A$)  (%_. True) (lt p a b)
  \end{lstlisting}
  This defines a context in which we have an abstract compare function \is{lt} for the abstract elements of type \is{'e}.
  It takes an extra parameter of type \is{'p} (e.g.\ the list $xs$), and forms a weak ordering\footnote{
    A weak ordering is induced by a mapping of the elements into a total ordering. It is the standard prerequisite for sorting algorithms in C++~\cite{Josu12}.
  }. Note that the strict compare function \is{lt} also induces a non-strict version \is{le p a b == \<not>lt p b a}.
  Moreover, we have a concrete implementation \is{lt$_\dagger$} of the compare function, wrt.\ the refinement assertions
  \is{par$_A$} for the parameter and \is{elem$_A$} for the elements.
